Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equations

### 2014-03-31 pituba/article/article.tex
Eternal rewritting...

\section{Introduction}
The so called material balance equations are widely employed in reservoir engineering and have become a classical tool in oil reservoir management. The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from phase equilibrium and balance of fluid production and volumetric expansion. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability, where most of the produced volume would come from and where pressure would present a more equal behavior. Such regions were said to contain an \textit{active oil}.

@@

Since black-oil and material balance equations differ only in the number of spatial dimensions and the existence of local pressure gradients, there must exist conditions under which material balance equations can be recovered from the black-oil approach. This existing link was briefly addressed in \cite{ertekin2001basic}, where both were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
\begin{subequations}
\begin{equation}
p_\alpha = p \, ,
\end{equation}
\begin{equation}
\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
\end{equation}
\end{subequations}
for $\alpha=o,g,w$ referring to oil, gas and water phases.

@@

We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature condition to the volume of this same phase in standard pressure and temperature, i.e.:

\begin{equation}
B_w = \frac{\rho_w^{ST}}{\bar{\rho_w}} \, ,
\end{equation}

\begin{equation}
B_o = \frac{\rho_o^{ST}}{\bar{Y_o} \bar{\rho_o}} \, ,
\end{equation}

\begin{equation}
B_g = \frac{\rho_g^{ST}}{\bar{\rho_g}} \, .
\end{equation}

@@

Performing time integration, balance equations for water and oil are:
\begin{equation}\label{eq: So5}
\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p \, ,
\end{equation}
\begin{equation}\label{eq: So5}
\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = \left(N-N_p\right) \, ,
\end{equation}
where $N=\bar{S_{o1}} V_{p1} / B_{o1}$ is the initial volume of oil measured in standard conditions.

@@

Assuming a constant isothermal compressibility to oil phase
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o
\end{equation}
and using hydrostatic equilibrium $dp_o = \rho_o g dz$, we obtain the expression for oil density variation with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

For the gas phase, we obtain its isothermal compressibility from gas state equation
\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial p_g}\right)_{T,Z} = \frac{\rho_g}{p_g} = \frac{\mathcal{M}}{ZRT}
\end{equation}

And again we use hydrostatic equilibrium to obtain gas density variation with depth:
\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \left(\frac{\mathcal{M}}{ Z R T}\right) g
\end{equation}

%In oil zone:
%\begin{equation}
%\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g
%\end{equation}

\begin{equation}
\left(\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1}
\end{equation}

\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1}
\end{equation}
If fluid densities do not vary appreciably with depth.

@@

An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used this library version $0.12$.

@@

It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects such as compositional gradients and thus must be used with care. For instance, it may be in strong disagreement with real reservoir behavior when gravity segregation is present.

>>>>

\section{Introduction}
The so called material balance equations are widely employed in reservoir engineering and have become a classical tool in oil reservoir management. The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from phase equilibrium and balance of fluid production and volumetric expansion. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a more equal behavior. Such regions were said to contain an \textit{active oil}.

@@

Since black-oil and material balance equations differ only in the number of spatial dimensions and the existence of local pressure gradients, there must exist conditions under which material balance equations can be recovered from the black-oil approach. This existing link was briefly addressed in \cite{ertekin2001basic}, where both approaches were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
\begin{subequations}
\begin{equation}
p_\alpha = p \, ,
\end{equation}
\begin{equation}
\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
\end{equation}
\end{subequations}
for $\alpha=o,g,w$ referring to oil, gas and water phases.

@@

We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of this same phase at standard pressure and temperature, i.e.:

\begin{equation}\label{eq: Bw}
B_w = \frac{\rho_w^{ST}}{\bar{\rho_w}} \, ,
\end{equation}

\begin{equation}\label{eq: Bo}
B_o = \frac{\rho_o^{ST}}{\bar{Y_o} \bar{\rho_o}} \, ,
\end{equation}

\begin{equation}\label{eq: Bg}
B_g = \frac{\rho_g^{ST}}{\bar{\rho_g}} \, .
\end{equation}

The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth, if effects of temperature and compositional gradients are neglected. If we put together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.

For showing this, we assume a constant isothermal compressibility to oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

Similarly for the gas phase, we obtain its isothermal compressibility from the gas state equation
\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial p_g}\right)_{T,Z} = \frac{1}{p_g} = \frac{\mathcal{M}}{\rho_g ZRT}
\end{equation}
and again we use hydrostatic equilibrium to obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradg}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

Substituting typical values in Eqs.~\eqref{eq: grado} and \eqref{eq: gradg}, we see that oil and gas densities are rather constant with depth:
\begin{equation}
\left(\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1} \, ,
\end{equation}

\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1} \, .
\end{equation}

The reference depth for computing fluid density (or formation volume factors) may be defined at gas-oil contact for a reservoir with a gas cap. At the fluid contact, both gas and oil densities may be computed using the corresponding local pressure. 
If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas zone. This is to avoid the effect of oil pressure gradient in the computation of gas density, i.e.,
\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g
\end{equation}
what is $\rho_o / \rho_g$ times greater than the pure effect of gas hydrostatic pressure on its density.

@@

Performing time integration, equations for water and oil are:
\begin{equation}\label{eq: So5}
\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p - W_i - W_e\, ,
\end{equation}
\begin{equation}\label{eq: So5}
\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = \left(N-N_p\right) \, ,
\end{equation}
where $N=\bar{S_{o1}} V_{p1} / B_{o1}$ is the initial volume of oil measured in standard conditions.

@@

@@

An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.

@@

It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects, e.g. compositional gradients, and must be used with care. For instance, it may be in strong disagreement with real reservoir behavior when gravity segregation is present.
